---
title: "S17305_DEAnalysis"
output: html_document
date: "2025-05-07"
---
```{r}
ataxiaInteractome <- read.csv("~/Documents/AtaxiaInteractome/ataxiaInteractome.csv")
S17305_alldata_original <- read.csv("~/Downloads/S17305_alldata_original.csv")
```

```{r}
library(DESeq2)
library(EnhancedVolcano)
library(ggplot2)
library(ggpubr)
```

# DESeq
```{r}
# run DESeq
dds <- DESeq(dds)
```

### 1. Ctrl8w_NaCl VS KO8w_NaCl
```{r}
dds_res1 <- results(dds,
                   alpha = 0.01, 
                   contrast=c("condition","KO8w_NaCl","Ctrl8w_NaCl"))
```

### 2. KO8w_AAV vs KO8w_NaCl
```{r}
dds_res2 <- results(dds,
                    alpha = 0.01, 
                    contrast=c("condition","KO8w_AAV","KO8w_NaCl"))
```

### 3. KO8w_AAV vs Ctrl8w_NaCl
```{r}
dds_res3 <- results(dds,
                    alpha = 0.01, 
                    contrast=c("condition","KO8w_AAV","Ctrl8w_NaCl"))
```

### 4. KO3w vs Ctrl3w
```{r}
dds_res4 <- results(dds,
                    alpha = 0.01,
                    contrast=c("condition","KO3w","Ctrl3w"))
```

# S17305_alldata_DESEQ2 File
```{r}
S17305_alldata_DESEQ2 <- data.frame(
  geneEnsemblID = S17305_alldata$Ensembl.Gene.ID,
  geneName = S17305_alldata_original$GeneName,
  
  
  lfc_KO8wNaCl_vs_Ctrl8wNaCl = dds_res1$log2FoldChange,
  pval_KO8wNaCl_vs_Ctrl8wNaCl = dds_res1$pvalue,
  padj_KO8wNaCl_vs_Ctrl8wNaCl = dds_res1$padj,
  
  
  lfc_KO8wAAV_vs_KO8wNaCl = dds_res2$log2FoldChange,
  pval_KO8wAAV_vs_KO8wNaCl = dds_res2$pvalue,
  padj_KO8wAAV_vs_KO8wNaCl = dds_res2$padj,
  
  
  lfc_KO8wAAV_vs_Ctrl8wNaCl = dds_res3$log2FoldChange,
  pval_KO8wAAV_vs_Ctrl8wNaCl = dds_res3$pvalue,
  padj_KO8wAAV_vs_Ctrl8wNaCl = dds_res3$padj,
  
  
  lfc_KO3w_vs_Ctrl3w = dds_res4$log2FoldChange,
  pval_KO3w_vs_Ctrl3w = dds_res4$pvalue,
  padj_KO3w_vs_Ctrl3w = dds_res4$padj)
```

```{r eval=FALSE, include=FALSE}
# export 
write.csv(S17305_alldata_DESEQ2, "/home/amani/Documents/S17305_DataAnalysis/S17305_alldata_DESEQ2.csv", row.names =  F, quote = F)
```

# DEG
### helper function
```{r}
get_deg <- function(res_table, lfc_threshold = 2, padj_threshold = 0.01) {
  deg <- res_table[
    !is.na(res_table$log2FoldChange) &
    !is.na(res_table$padj) &
    abs(res_table$log2FoldChange) > lfc_threshold &
    res_table$padj < padj_threshold,
  ]
  
  names_deg <- rownames(deg)
  list(deg_table = deg, deg_names = names_deg)
}
```

### 1. Ctrl8w_NaCl VS KO8w_NaCl
```{r}
results1 <- get_deg(dds_res1)
deg1 <- results1$deg_table #DEG subset of DESeq 
names_deg1 <- results1$deg_names #DEG names
length(names_deg1)
```

### 2. KO8w_AAV vs KO8w_NaCl
```{r}
results2 <- get_deg(dds_res2)
deg2 <- results2$deg_table #DEG subset of DESeq 
names_deg2 <- results2$deg_names #DEG names
length(names_deg2)
```

### 3. KO8w_AAV vs Ctrl8w_NaCl
```{r}
results3 <- get_deg(dds_res3)
deg3 <- results3$deg_table #DEG subset of DESeq 
names_deg3 <- results3$deg_names #DEG names
length(names_deg3)
```

### 4. KO3w vs Ctrl3w
```{r}
results4 <- get_deg(dds_res4)
deg4 <- results4$deg_table #DEG subset of DESeq 
names_deg4 <- results4$deg_names #DEG names
length(names_deg4)
```

# Table
```{r}
deg_table <- data.frame(
  Comparison = c("KO8w_NaCl vs WT8w_NaCl", "KO8w_AAV vs KO8w_NaCl", "KO8w_AAV vs WT8w_NaCl", "KO3w vs WT3w"),
  
  DEG_Count = c(nrow(deg1), nrow(deg2), nrow(deg3),
                nrow(deg4)), 
  
  Up_DEG = c((sum(deg1$log2FoldChange > 2)), (sum(deg2$log2FoldChange > 2)),
             (sum(deg3$log2FoldChange > 2)), (sum(deg4$log2FoldChange > 2))), 
  
  Down_DEG = c((sum(deg1$log2FoldChange < -2)), (sum(deg2$log2FoldChange < -2)),
             (sum(deg3$log2FoldChange < -2)), (sum(deg4$log2FoldChange < -2)))
   )
```

# DEG & FRDA lists 
### Get Gene Names (from ensembl ID)
```{r}
### 1. Ctrl8w_NaCl VS KO8w_NaCl
identity1 <- S17305_alldata_original[S17305_alldata_original$EnsemblGeneID %in% names_deg1, ]$GeneName

### 2. KO8w_AAV vs KO8w_NaCl
identity2 <- S17305_alldata_original[S17305_alldata_original$EnsemblGeneID %in% names_deg2, ]$GeneName

### 3. KO8w_AAV vs Ctrl8w_NaCl
identity3 <- S17305_alldata_original[S17305_alldata_original$EnsemblGeneID %in% names_deg3, ]$GeneName

### 4. KO3w vs Ctrl3w
identity4 <- S17305_alldata_original[S17305_alldata_original$EnsemblGeneID %in% names_deg4, ]$GeneName
```


## FRDA gene List 
The FRDA gene list contains genes that were identified from 4 differential expression analyses:
- WT_8w_NaCl vs KO_8w_NaCl (DEG1)
- KO_8w_NaCl vs KO_8w_AAV (DEG2)
- WT_8w_NaCl vs KO_8w_AAV (DEG3)
- WT_3w vs KO_3w (DEG4)

For each comparison, the DEGs were extracted and compared with the interactome. The final list includes only the unique genes that appeared in at least one of the analyses. Genes that overlap between DEG1 and DEG2 were filtered (they correspond to the genes whose level return to normal after treatment, thus not interesting for the ataxia interactome).

### list creation 
```{r}
# list of all DEG 
DEGList <-unique(c(identity1, identity2, identity3, identity4))
cat("# DEG:", length(DEGList), "\n")

# list of DEGs not in the ataxia interactome
FRDAList <- setdiff(DEGList, ataxiaInteractome$Identity) 
cat("# DEG not in the Ataxia Interactome:", length(FRDAList), "\n")
```

### filter genes whose level return to normal after treatmemt
```{r}
# get overlapping genes between DEG 1 & DEG 2 
DEG1_2_overlap <- intersect(identity2, identity1)

# remove overlapping gene names from newDEGList
FRDAList <- setdiff(FRDAList, DEG1_2_overlap)
cat("# DEG not in the Ataxia Interactome (after filtering):", length(FRDAList), "\n")
```

### add lfc & p-adj to FRDA List 
```{r}
FRDAList <- data.frame(gene = FRDAList)

# get Ensembl ID
FRDAList$Ensembl_ID <- S17305_alldata_original$EnsemblGeneID[match(FRDAList$gene, S17305_alldata_original$GeneName)]

# list of DEG dataframes
deg_list <- list(deg1, deg2, deg3, deg4)

# helper function : extract LFC & p-adj
extract_vals <- function(deg_df, column_name) {
  return(deg_df[[column_name]][match(FRDAList$Ensembl_ID, rownames(deg_df))])
}

# loop
for (i in 1:length(deg_list)) {
  FRDAList[[paste0("lfc_DEG", i)]] <- extract_vals(deg_list[[i]], "log2FoldChange")
  FRDAList[[paste0("padj_DEG", i)]] <- extract_vals(deg_list[[i]], "padj")
}


```

### add lfc & p-adj to DEG List 
```{r}
DEGList <- data.frame(gene = DEGList)

# get Ensembl ID
DEGList$Ensembl_ID <- S17305_alldata_original$EnsemblGeneID[match(DEGList$gene, S17305_alldata_original$GeneName)]

# list of DEG dataframes
deg_list <- list(deg1, deg2, deg3, deg4)

# helper function : extract LFC & p-adj
extract_vals <- function(deg_df, column_name) {
  return(deg_df[[column_name]][match(DEGList$Ensembl_ID, rownames(deg_df))])
}

# loop
for (i in 1:length(deg_list)) {
  DEGList[[paste0("lfc_DEG", i)]] <- extract_vals(deg_list[[i]], "log2FoldChange")
  DEGList[[paste0("padj_DEG", i)]] <- extract_vals(deg_list[[i]], "padj")
}


```

### export
```{r eval=FALSE, include=FALSE}
# export FRDA List (only DEG that are not in the interactome)
write.csv(FRDAList, file = "/home/amani/Documents/S17305_DataAnalysis/FRDAList.csv", row.names = F, quote = F)

# export the DEG List (all DEG from the 4 DE Analyses)
write.csv(DEGList, file = "/home/amani/Documents/S17305_DataAnalysis/DEGList.csv", row.names = F, quote = F)
```

# MA-Plots
```{r eval=FALSE, include=FALSE}
# Ctrl8w_NaCl VS KO8w_NaCl
plotMA(dds_res1, alpha = 0.01,
       ylim = c(-2,2), 
       colSig = "#fc8d62")

# KO8w_AAV vs KO8w_NaCl
plotMA(dds_res2, alpha = 0.01,
       ylim = c(-2,2), 
       colSig = "#fc8d62")

# KO8w_AAV vs Ctrl8w_NaCl
plotMA(dds_res3, alpha = 0.01,
       ylim = c(-2,2), 
       colSig = "#fc8d62")

# KO3w vs Ctrl3w
plotMA(dds_res4, alpha = 0.01,
       ylim = c(-2,2), 
       colSig = "#fc8d62")
```

# Volcano-Plot 
### helper function 
```{r}
plotVolcano_func <- 
  function(res, subtitle, comparison, fc_cutoff = 2, p_cutoff = 0.01) {
  keyvals <- ifelse(
    res$log2FoldChange < -fc_cutoff & res$padj < p_cutoff, '#64c3a3',
    ifelse(res$log2FoldChange > fc_cutoff & res$padj < p_cutoff, '#fa8e64', 'darkgrey'))
  keyvals[is.na(keyvals)] <- 'darkgrey'
  names(keyvals)[keyvals == '#fa8e64'] <- 'Up-regulated DEGs'
  names(keyvals)[keyvals == 'darkgrey'] <- 'Non-significant DEGs'
  names(keyvals)[keyvals == '#64c3a3'] <- 'Down-regulated DEGs'

  EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = "",
    xlab = bquote(~log[2]~ 'FC'),
    axisLabSize = 18,
    pCutoff = p_cutoff,
    FCcutoff = fc_cutoff,
    colCustom = keyvals,
    pointSize = 1,
    labSize = 12,
    colAlpha = 1,
    legendPosition = 'top',
    legendLabSize = 18,
    legendIconSize = 5,
    title = NULL,
    caption = NULL,
    titleLabSize = 0,
    subtitle = subtitle, 
    subtitleLabSize = 18
    ) +
    annotate(geom = "text", x=5, y=40, label = deg_table[deg_table$Comparison == comparison, "Up_DEG"], size = 10
) +
    annotate(geom = "text", x= -5, y=40, label = deg_table[deg_table$Comparison == comparison, "Down_DEG"], size = 10)
  
}
```

### plots
```{r}
volcano_plot1 <- plotVolcano_func(dds_res1, "1", "KO8w_NaCl vs WT8w_NaCl")+ theme(legend.position = "top", legend.justification = "center")
volcano_plot2 <- plotVolcano_func(dds_res2, "2", "KO8w_AAV vs KO8w_NaCl")+ theme(legend.position = "top", legend.justification = "center")
volcano_plot3 <- plotVolcano_func(dds_res3, "3", "KO8w_AAV vs WT8w_NaCl")+ theme(legend.position = "top", legend.justification = "center")
volcano_plot4 <- plotVolcano_func(dds_res4, "4", "KO3w vs WT3w")+ theme(legend.position = "top", legend.justification = "center")
```


# Barplots
```{r eval=FALSE, include=FALSE}
# import Friedreich Annotated DEG List
DEGList <- read.csv("~/Documents/S17305_DataAnalysis/DEGList.csv")
```

### helper function
```{r}
degBarPlot <- function(lfc, title) {
  
  #add DE label
  deg_plot <- DEGList
  deg_plot$DEG <- ifelse(deg_plot[[lfc]] > 0, "Up-regulated", "Down-regulated")
  
  #select up-regulated genes 
  up_genes <- deg_plot[deg_plot[[lfc]] > 0, ]
  up_genes <- up_genes[order(up_genes[[lfc]], decreasing = TRUE), ][1:10, ]
  
  #select down-regulated genes
  down_genes <- deg_plot[deg_plot[[lfc]] < 0, ]
  down_genes <- down_genes[order(down_genes[[lfc]]), ][1:10, ]
  
  top_genes <- rbind(up_genes, down_genes)
  
  #plot
  ggbarplot(top_genes, x = "gene", y = lfc,
            fill = "DEG",           
            color = "white",            
            palette = "Set2",            
            sort.val = "desc",          
            sort.by.groups = FALSE,     
            x.text.angle = 90,          
            title = title,
            legend.title = "",
            rotate = TRUE,
            ggtheme = theme_minimal()) + 
    labs(
    y = expression(log[2](FC)),
    x= "Gene") + 
    theme(
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18), 
      axis.text = element_text(size = 16),
      legend.title = element_text(size = 18), 
      legend.text = element_text(size = 18))
}
```

### plots
```{r}
deg_plot1 <- degBarPlot("lfc_DEG1", "5")
deg_plot2 <- degBarPlot("lfc_DEG2", "6")
deg_plot3 <- degBarPlot("lfc_DEG3", "7")
deg_plot4 <- degBarPlot("lfc_DEG4", "8")
```

# Final figure for report 
```{r}
# volcano plots
volcano_panel <- ggarrange(
  volcano_plot1, volcano_plot2,
  volcano_plot3, volcano_plot4,
  ncol = 1, nrow = 4,
  common.legend = TRUE,
  legend = "top")

volcano_panel <- annotate_figure(volcano_panel,
               top = text_grob("A", face = "bold", size = 22))

#barplots
barplot_panel <- ggarrange(
  deg_plot1, deg_plot2,
  deg_plot3, deg_plot4,
  ncol = 1, nrow = 4,
  common.legend = TRUE,
  legend = "top")

barplot_panel <- annotate_figure(barplot_panel,
               top = text_grob("B", face = "bold", size = 22))
# final fig
VolcanoBarPlots <- ggarrange(
  volcano_panel,
  barplot_panel,
  ncol = 2)
```

```{r eval=FALSE, include=FALSE}
ggsave("VolcanoBarPlots.pdf", VolcanoBarPlots,
       width = 18, height = 22, units = "in", dpi = 320, bg = "white")
```




